Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.tex
edits and updates to 2.2

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks as big as 12
\citep{midas2020,biggerstaff2014,namee2018}.

Estimator(s) for \(R_0\) are dependent on the epidemic modeling
framework, which consists of which states an individual can occupy
(e.g.~susceptible, infectious, recovered) and a description of how
individuals move from one state to the next over time (see
\cite{hethcote1994}). A common epidemic modeling framework is the SIR
model, originally introduced by \citet{kermack1927}. Transitions from
one state to the next are defined by a series of ordinary differential
equations, where \(N\) is the (fixed) total number individuals,
\(\beta\) is the rate of infection, and \(\gamma\) is rate of recovery,
\begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align} From this,
\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
estimated infection rate compared to the estimated recovery rate.

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15.

@@

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig.
\ref{fig:different-scales-tern}, the points seem to overlap and form the
same trajectory. In other words, we can see there is something
fundamentally linking these two different epidemics, and this
fundamental link turns out to be \(R_0\).

More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode}
) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\) and
\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

@@

But time-invariant analysis is also useful for epidemic models with more
than three states. The constraints in three dimensions that are met with
the SIR model (there are three epidemic states and for each time step,
the sum of total number of individuals in each state is the population
size at that time\$) actually represents a space of 3d simplices, and
the ternary plot specifically represents these simplices, after scaling
to examine such values as proportions between 0 and 1 (ternary plots are
known as a 3d unit simplex due to it's scaling). This same scaling for
larger models (i.e.~with more states) can be done onto different
simplexes. In this package we present tools to help compare models
(mostly through simulations). s tool compares these objects after
projecting them into a one-dimension-fewer space through the simplexical
structure of the data.

In \pkg{EpiCompare} we allow for the comparison epidemics in these
higher dimensional spaces by first projecting onto these simplexes. Even
though higher dimensional models may not be able to visualized, we
provide multiple tools to aid in the comparison of models and epidemics.
The first of which uses multiple simulations under specific model
parameters to assess the bairabilty of the model fit. In
\pkg{EpiCompare} we provide ways to create prediction regions for a true
epidemic under the model assumptions using these simulations. These
regions require representing multi-dimensional structures for functions
to completely contain epidemics, and treat these simulations and
epidmeics as filamental objects. We extend off of papers like
\citet{Dalmasso2019a} to create these bands. These high dimensional
bands allow the user to assess if the true epidemic is within the band
(there-by assessing the model's representation of the epidemic), and
also compare who different models are from each other through distances
that compare sets. We recommend using the Hausdorff distance to compare
such sets as it captures how much bigger the sets would have to expand
to cover each other, and is defined mathematically as \[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

>>>>

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks as large as 12
\citep{midas2020,biggerstaff2014,namee2018}.

Estimators for \(R_0\) are dependent on the epidemic modeling framework,
which consists of which states an individual can occupy
(e.g.~susceptible, infectious, recovered) and a description of how
individuals move from one state to the next over time (see
\cite{hethcote1994}). A common epidemic modeling framework is the SIR
model, originally introduced by \citet{kermack1927}. Transitions from
one state to the next are defined by a series of ordinary differential
equations, where \(N\) is the (fixed) total number individuals,
\(\beta\) is the rate of infection, and \(\gamma\) is rate of recovery,
\begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align} From this,
\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
estimated infection rate compared to the estimated recovery rate.

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over the first 15 days. Both
epidemics have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15.

@@

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig.
\ref{fig:different-scales-tern}, the points seem to overlap and form the
same trajectory. Now it seems to be that Model 2 is following the same
trajectory as Model 1 but is not as far along in the infection process.
We can see there is something fundamentally linking these two different
epidemics, and this fundamental link turns out to be \(R_0\).

More formally, let two Kermack and McKendrick SIR models(see
Eq.\eqref{eq:sir-ode} ) be denoted \((S_1(t), I_1(t), R_1(t))\) and
\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

@@

Moreover, time-invariant analysis is also useful for epidemic models
with more than three epidemic states. The constraints in three
dimensions that are met with the SIR model (there are three epidemic
states and for each time step, the sum of total number of individuals in
each state is the population size at that time) represents a space of 3d
simplices, and the ternary plot specifically represents these simplices,
after scaling the values in each state as proportions of the total
population at that time (ternary plots are known as a 3d unit simplex
due to it's scaling). This same scaling for larger models (i.e.~with
more states) can be done onto different simplexes. In this package we
present tools to help compare models (mostly through simulations).
\pkg{EpiCompare}'s tools permits comparisons of \(d\)-dimensional
objects by projecting them into (\(d-1\))-dimensional space.

Even though higher dimensional models may not be able to visualized
simply, we provide a number of tools to aid in the comparison of models
and epidemics. The first of which uses multiple simulations under
specific model parameters to assess the \textcolor{orange}{variability?}
bairabilty of the model fit. In \pkg{EpiCompare}, we provide ways to
create prediction regions for a `true' epidemic under a fully specified
model. These regions require representing multi-dimensional structures
for functions to completely contain epidemics. \pkg{EpiCompare} treats
these simulations and epidemics as \textit{filamental} objects --
\textcolor{orange}{need a description for filamental, Ben feel free to change}
where not only the outbreak trajectory points are important but also
their form which can be considered a function.
\textcolor{orange}{could be useful to have a picture/example here? like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance? I'll sketch up an idea unless you have something in your proposal already}

We extend off of papers like \citet{Dalmasso2019a} to create these
bands. These high dimensional bands allow the user to assess if the true
epidemic is within the band (thereby assessing the model's
representation of the epidemic). With these bands, we can compare
\textit{models} to \textit{models} by assessing the distance between the
two bands of model representation. We recommend using the Hausdorff
distance to compare two bands to one another, as this distance captures
how much bigger the sets would have to expand to cover each other. The
Hausdorff distance is defined mathematically as \[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]
